From heartbeat to data -- Using wearable fitness trackers as an affordable approach to assess teacher stress

### 2024-09-10 02_paper/02_study/01_fitbit paper/Submission/FitBit_submission.tex
Format_main

More recent studies have begun using wrist-worn fitness trackers to
investigate HR trends in instructional settings
\citep{Darnell2019, chalmers2021}. \citet{Darnell2019} measured the HR
of 15 medical college students listening to lectures, using wrist-worn
devices. The analysis revealed a constant decrease in HR throughout the
lecture, with HR peaks during more interactive learning phases.
\citet{chalmers2021} used HR data from a fitness tracker to identify
physiological changes during stress-inducing tasks (i.e., the Trier
Social Stress Test; \citet{kirschbaum1993trier}{]}. Average HR increased
significantly from the resting to the stress inducing phases. Even
though the participants of these previous studies
\citep{Darnell2019, chalmers2021} were not teachers but learners, the
results are relevant for studying teacher stress as they demonstrated
that HR can be effectively recorded using fitness trackers over the
course of a whole learning unit, and HR changes are in line with the
occurrence of activating or stress-inducing tasks.

@@

According to \citet{kyriacou1978}, teacher stress can be defined as a
negative affective response, typically accompanied by physiological
changes such as increased HR, triggered by job-related demands, and
perceived as threatening to one's self-esteem or well-being. Coping
mechanisms help to reduce the perceived threat. Kyriacou's definition of
teacher stress (see \citet{kyriacou1978} and, for a more recent
adaptation, \citet{van2006stress}) is based on the transactional stress
model by Lazarus and colleagues {[}lazarus1966psychological;
\citet{lazarus1990theory}{]}, which highlights the interaction between
an individual and the environment, whereby stress refers to a person's
subjective reaction to an event (a stressor) that exceeds the person's
adaptive resources.

@@

To measure teachers' HR, we used the wrist-based fitness tracker Fitbit®
Charge 4. In line with the manufacturer's instructions \citep{fitbitnd},
the device was attached to the participants' nondominant hand, a
finger's width above the wrist bone. The tracker works by flashing green
LEDs hundreds of times per second, using light-sensitive photodiodes to
catch the reflected light, to calculate the volume changes in the
capillaries. From this, the tracker calculated the heart beats per
minute. HR measurements were generated at least every 15 seconds . The
raw data contained the estimated HR in BPM for each time stamp. To
account for individual differences in the baseline HR, we also
calculated z-standardized HR values based on individual means, i.e., at
the subject level of \(n\) = 81 participants (standardized HR).

@@

@@

\begin{table}[ht]
    \centering
    \begin{tabularx}{\textwidth}{lXXXXX}
        \toprule
        Interval & \textit{M HR} & \textit{SD HR} & Min & Max \\
        \midrule
        Overall Course of 2h & 90.09/0.04\tablefootnote{Please note that standardized M and SD of the overall course were not exactly 0 and 1 due to rounding differences} & 15.76/0.991 & 51/-4.03 & 164/4.56 \\
        Pre-teaching interval ($I_1$) & 96.28/0.48 & 14.11/0.88 & 56/-3.56 & 139/3.24 \\
        Teaching interval ($I_2$) & 100.80/0.85 & 16.23/0.77 & 63/-2.18 & 164/4.37 \\
        Post-teaching interval ($I_3$) & 93.61/0.27 & 14.01/0.76 & 60/-2.17 & 150/3.06 \\
        Interview interval ($I_4$) & 82.32/-0.72 & 11.85/0.74 & 51/-2.51 & 132/4.39 \\
        End interval $I_5$) & 77.95/-1.07 & 11.14/0.57 & 50\tablefootnote{Deviations of the minimum values in the overall course vs. the end interval ($I_5$) are due to data of a few participants who
        needed more than two hours to finish the study.}/-2.68 & 120/2.96 \\
        \bottomrule
    \end{tabularx}
    \caption{Mean HR (M), standard deviations HR (SD), and range of teachers’ HR over the course of the entire study and the five intervals (unstandardized in BPM/z-standardized).}
    \label{tab_1}
\end{table}

@@

@@

@@

\begin{table}[ht]
    \centering
    \begin{tabularx}{\textwidth}{lXXXXX}
        \toprule
        Interval & n\tablefootnote{All measurement points per interval for all participants. Note that the variation in $n$ stems from the variation in the number of collected data points by the fitness tracker.} & \multicolumn{2}{c}{\textit{M (SD)}} & \multicolumn{2}{c}{$p$} \\
        & & Intercept & Slope & Intercept & Slope \\
        \midrule
        (1) Pre-teaching interval & 6896 & 0.052  & 0.085*  & .57 & $<$ .05 \\
        & & (0.820) & (0.133) & & \\
        (2) Teaching interval & 7150 & 1.025* & -0.039* & $<$ .05 & $<$ .05 \\
        & & (0.690) & (0.108) & & \\
        (3) Post-teaching interval & 6664 & 0.549* & -0.060* & $<$ .05 & $<$ .05 \\
        & & (0.547) & (0.101) & & \\
        (4) Interview interval & 6287 & -0.617* & -0.022 & $<$ .05 & .006 \\
        & & (0.614) & (0.070) & & \\
        (5) End interval & 5990 & -1.004* & -0.012 & $<$ .05 & .14 \\
        & & (0.500) & (0.074) & & \\
        \bottomrule \\
        Note. * $p < .05$
    \end{tabularx}
    \caption{Descriptive statistics (n, M, SD) for the mean intercepts and the mean slopes for the five intervals.}
    \label{tab_2}
\end{table}

@@

>>>>

More recent studies have begun using wrist-worn fitness trackers to
investigate HR trends in instructional settings
\citep{Darnell2019, chalmers2021}. \citet{Darnell2019} measured the HR
of 15 medical college students listening to lectures, using wrist-worn
devices. The analysis revealed a constant decrease in HR throughout the
lecture, with HR peaks during more interactive learning phases.
\citet{chalmers2021} used HR data from a fitness tracker to identify
physiological changes during stress-inducing tasks (i.e., the Trier
Social Stress Test; \citet{kirschbaum1993trier}). Average HR increased
significantly from the resting to the stress inducing phases. Even
though the participants of these previous studies
\citep{Darnell2019, chalmers2021} were not teachers but learners, the
results are relevant for studying teacher stress as they demonstrated
that HR can be effectively recorded using fitness trackers over the
course of a whole learning unit, and HR changes are in line with the
occurrence of activating or stress-inducing tasks.

@@

According to \citet{kyriacou1978}, teacher stress can be defined as a
negative affective response, typically accompanied by physiological
changes such as increased HR, triggered by job-related demands, and
perceived as threatening to one's self-esteem or well-being. Coping
mechanisms help to reduce the perceived threat. Kyriacou's definition of
teacher stress (see \citet{kyriacou1978} and, for a more recent
adaptation, \citet{van2006stress}) is based on the transactional stress
model by Lazarus and colleagues
\citep{lazarus1966psychological, lazarus1990theory}, which highlights
the interaction between an individual and the environment, whereby
stress refers to a person's subjective reaction to an event (a stressor)
that exceeds the person's adaptive resources.

@@

To measure teachers' HR, we used the wrist-based fitness tracker Fitbit®
Charge 4. In line with the manufacturer's instructions \citep{fitbitnd},
the device was attached to the participants' nondominant hand, a
finger's width above the wrist bone. The tracker works by flashing green
LEDs hundreds of times per second, using light-sensitive photodiodes to
catch the reflected light, to calculate the volume changes in the
capillaries. From this, the tracker calculated the heart beats per
minute. HR measurements were generated at least every 15
seconds\footnote{The fluctuations in the number of seconds in which the
  HR was measured are due to the participants' movements, meaning that
  the device could not measure the HR every second.}. The raw data
contained the estimated HR in BPM for each time stamp. To account for
individual differences in the baseline HR, we also calculated
z-standardized HR values based on individual means, i.e., at the subject
level of \(n\) = 81 participants (standardized HR).

@@

\subsubsection{Teaching Experience}\label{teaching-experience}

Participants' teaching experience was assessed as a part of their
sociodemographic data. Participants stated their work experience in
years.

\subsubsection{Subjective appraisal of the classroom events and coping
processes}\label{subjective-appraisal-of-the-classroom-events-and-coping-processes}

The subjective disruption and confidence appraisals were assessed during
the SRI on an 11-point rating scale, ranging from 0 (not at all
disrupting/confident) to 10 (extremely disrupting/confident). Ratings
were averaged across the nine classroom events for each participant, as
we were interested in the general stressfulness of the \emph{teaching
phase} for each participant.

\subsection{Data analysis}\label{data-analysis}

We conducted all analyses with R \citep{RStudio2020}. Graphics were
created using ggplot2 \citep{ggplot2}.

To enable visual inspection of HR trends, we displayed smoothed teacher
HR over the course of the recording\footnote{The curve was smoothed
  using the geom\_smooth() function from the ggplot2 package in R
  \citep{ggplot2} based on the smoothing method LOESS (Locally Estimated
  Scatterplot Smoothing). This method fits a polynomial surface
  determined by one or more numerical predictors, using local fitting.}.
We visually compared unstandardized and standardized HR trends over the
two-hour recording period\footnote{Note that the study exceeded the
  planned duration of two hours for a few participants. To avoid
  distortions when mapping the HR over the course of the study (see Fig.
  3), the endpoint was set at two hours for all participants, even
  though data from later time points was used in the \emph{end interval}
  for a few participants.}. For all further analyses, we used
standardized rather than unstandardized HR values.

We averaged each person's standardized HR over each of the five selected
intervals\footnote{We used the mean standardized HR instead of the mean
  intercept as we wanted to explain the mean HR of the entire intervals
  and not the HR at the very beginning of the interval (\(x\) = 0).},
resulting in one measure per person per interval. To test Hypothesis 1a,
we initially conducted a one-way ANOVA with repeated measures as an
omnibus test and then tested the mean differences between the
\emph{teaching interval} (\(I_2\)) and the other four intervals by
planned contrasts and inspection of effect size \(d\)
\citep{cohen1988statistical}.

For testing Hypothesis 1b, concerning HR changes within each interval,
we first conducted a linear estimation of the increase or decrease in
standardized HR values over time for each participant. To this end, we
used fixed intercept fixed slope regression models
\citep{gelman2006data} for each interval to estimate intercepts and
linear slopes for each individual, which were then averaged across
individuals.\footnote{Although this procedure does not account for
  nonmonotonic progressions in individual HR, a graphical evaluation
  revealed that the linear estimates corresponded well to the majority
  of the cases (see XX in the supplementary material).} We tested
Hypothesis 1b based on the unstandardized estimates of mean slopes (one
estimate per participant per interval).

Addressing our second research goal, we ran linear regression analysis
with teaching experience and subjective appraisals as predictors. To
test Hypothesis 2a, we examined the effect of teaching experience on
participants' HR levels (i.e., mean standardized HR) for each of the
five intervals, using linear regression models with teaching experience
as the sole predictor. To test Hypotheses 2b and 2c, we separately
augmented the model by either teachers' disruption appraisal (Hypothesis
2b) or confidence appraisal (Hypothesis 2c) as predictors, while
controlling for teaching experience. To test Hypothesis 2d, we examined
the effects of all three predictors in one regression model.
Furthermore, we repeated these steps to explore the effects of teaching
experience and subjective appraisals on \emph{changes} in teachers' HR
(i.e., mean slopes). Please note: HR levels and changes were not
regressed on the disruption and confidence appraisals in the
\emph{pre-teaching interval} (\(I_1\)), because the appraised classroom
events had not yet taken place in that phase.

\section{Results}\label{results}

\subsection{Mapping teachers' HR over the course of the study
phases}\label{mapping-teachers-hr-over-the-course-of-the-study-phases}

Means, standard deviations, and range of teachers' unstandardized and
standardized HR for the entire study period, and for the five intervals,
are shown in Table 1. Fig. 3 a. and b. display the unstandardized and
standardized HR trends, respectively, over the course of the entire
study period. HR initially increased, peaked, and then decreased, with
the unstandardized and standardized HR graphs showing high similarity.
Thus, for all further analyses, we used participants' standardized HR
values.

@@

\begin{table}[ht]
    \centering
    \begin{tabularx}{\textwidth}{lXXXXX}
        \toprule
        Interval & \textit{M HR} & \textit{SD HR} & Min & Max \\
        \midrule
        Overall Course of 2h & 90.09/0.04\footnotemark[9] & 15.76/0.991 & 51/-4.03 & 164/4.56 \\
        Pre-teaching interval ($I_1$) & 96.28/0.48 & 14.11/0.88 & 56/-3.56 & 139/3.24 \\
        Teaching interval ($I_2$) & 100.80/0.85 & 16.23/0.77 & 63/-2.18 & 164/4.37 \\
        Post-teaching interval ($I_3$) & 93.61/0.27 & 14.01/0.76 & 60/-2.17 & 150/3.06 \\
        Interview interval ($I_4$) & 82.32/-0.72 & 11.85/0.74 & 51/-2.51 & 132/4.39 \\
        End interval ($I_5$) & 77.95/-1.07 & 11.14/0.57 & 50\footnotemark[10]/-2.68 & 120/2.96 \\
        \bottomrule
    \end{tabularx}
    \caption{Mean HR (M), standard deviations HR (SD), and range of teachers’ HR over the course of the entire study and the five intervals (unstandardized in BPM/z-standardized).}
    \label{tab_1}

\end{table}
 \footnotetext[9]{Please note that standardized M and SD of the overall course were not exactly 0 and 1 due to rounding differences.}
 \footnotetext[10]{Deviations of the minimum values in the overall course vs. the end interval ($I_5$) are due to data of a few participants who needed more than two hours to finish the study.}

@@

\newpage

Figure 4 shows the distribution of teachers' mean standardized HR for
the five intervals. Repeated measures ANOVA revealed significant
differences in mean standardized HR between intervals,
\(F(4, 400) = 260.62\), \(p < .05\), \(d = 1.60\) (large effect).
Planned contrasts indicated that, as hypothesized (Hypothesis 1a), mean
standardized HR was significantly higher in the \emph{teaching interval}
(\(I_2\)) than in all other intervals, specifically, the
\emph{pre-teaching interval} (\(I_1\); \(t(400) = -10.08\), \(p < .05\),
\(d = 1.034\); large effect), the post-teaching interval (\(I_3\);
\(t(400) = -6.94\), \(p < .05\), \(d = 1.37\); large effect), the
interview interval (\(I_4\); \(t(400) = 15.00\), \(p < .05\),
\(d = 3.29\); large effect), and the end interval (\(I_5\));
\(t(400) = 22.54\), \(p < .05\), \(d = 4.64\); large effect).

@@

Next, we examined HR changes (i.e., mean slopes) within each interval to
test the hypothesis that HR increased during the \emph{pre-teaching
phase} and decreased during all other phases (Hypothesis 1b). The mean
intercepts and mean slopes, complemented by their standard deviations
for each interval, are shown in Table 2. The mean slope of the
\emph{pre-teaching interval} (\(I_1\)) was significantly positive,
indicating an increase in HR, as hypothesized. Further, the mean slopes
of the \emph{teaching interval} (\(I_2\)), \emph{post-teaching interval}
(\(I_3\)), and \emph{interview interval} (\(I_4\)) were significantly
negative, indicating a decrease in HR. For the last interval, the
\emph{end interval} (\(I_5\)), the mean slope was negative, but did not
differ significantly from zero.

@@

\begin{table}[ht]
    \centering
    \begin{tabularx}{\textwidth}{lXXXXXX}
        \toprule
        Interval  & \multicolumn{2}{c}{M (SD)} & \multicolumn{2}{c}{$p$} \\ & Intercept & Slope & Intercept & Slope \\
        \midrule
        (1) Pre-teaching interval & 0.052 (0.820) & 0.085 (0.133) & .57 & $< .05$ \\
        (2) Teaching interval & 1.025 (0.690) & -0.039 (0.108) & $< .05$ & $< .05$ \\
        (3) Post-teaching interval & 0.549 (0.547) & -0.060 (0.101) & $< .05$ & $< .05$ \\
        (4) Interview interval & -0.617 (0.614) & -0.022 (0.070) & $< .05$ & $< .05$ \\
        (5) End interval & -1.004 (0.500) & -0.012 (0.074) & $< .05$ & .14 \\
        \bottomrule
    \end{tabularx}
    \caption{Analysis (\textit{M, SD, p-values}) for the mean intercepts and the mean slopes for the five intervals }
    \label{tab_2}
\end{table}

\subsection{Predicting mean standardized HR and mean
slopes}\label{predicting-mean-standardized-hr-and-mean-slopes}

Table 3 shows the raw correlations among mean standardized HR/mean
slopes (see Table 2 for means and standard deviations), teaching
experience (\(M = 5.64\), \(SD = 9.46\)), disruption appraisal
(\(M = 5.19\), \(SD = 2.87\)), and confidence appraisal (\(M = 7.81\),
\(SD = 1.97\)). With a few notable exceptions, correlations with HR
measures were mostly very small and statistically non-significant (Table
3). Correlations between teaching experience and appraisals (not shown
in Table 3) were substantial: more experienced teachers gave lower
disruption appraisals (\(r = -.36\)), and higher confidence appraisals
(\(r = .44\)). Moreover, the two appraisal variables were negatively
correlated (\(r = -.37\)).

Table 4 shows the results of the regression analyses. Teaching
experience significantly predicted mean standardized HR only in the
\emph{interview interval} (Table 4, Interview interval, Model 1),
indicating a higher mean standardized HR for teachers with more teaching
experience. This relationship is, in fact, in the opposite direction as
predicted by Hypothesis 2a. Neither adding disruption appraisal
(\textbf{Hypothesis 2b}) nor adding confidence appraisal
(\textbf{Hypothesis 2c}) increased the amount of explained variance to a
statistically significant extent.

When considering the effects of the three predictors in concert
(\textbf{Hypothesis 2d}), mean standardized HR was significantly
predicted only by disruption appraisal, and only in the
\emph{post-teaching interval} (Table 4, Post-teaching interval, Model
4), indicating a higher mean standardized HR for teachers who felt more
disrupted by the classroom events, when controlling for the other
variables.

Concerning the explorative investigation of the effects of teaching
experience and subjective appraisals on \emph{changes} (i.e., mean
slopes) in teachers' HR, teaching experience significantly predicted the
mean slope in the \emph{pre-teaching interval} (Table 4, Pre-teaching
interval, Model 1), indicating a less steep HR increase in teachers with
more teaching experience. For all other intervals, no variable had
significant predictive value.

@@

\section{Discussion}\label{discussion}

\subsection{Key findings}\label{key-findings}

Overall, our findings indicate that wrist-worn fitness trackers are a
useful tool for tracking teachers' HR and identifying stressful periods
during teaching. Using HR data from a commercially available and
relatively low-cost Fitbit® fitness tracker, we were able to map
teachers' HR before, during, and after a stressful micro-teaching unit,
with HR increasing in preparation for teaching, peaking during the
teaching phase, and decreasing afterward.

These findings are in line with prior studies showing that teachers' HR
varies depending on their activities and encountered stressors with
increases during phases where teachers are in an exposed position
\citep{sperka1995, scheuch1997psychophysische, donker2018, junker2021},
as well as with findings showing how HR changes align with activating
events and stress-inducing tasks \citep{Darnell2019, chalmers2021}.

Building on the model of teacher stress \citep[see Fig.
2]{kyriacou1978}, we had hypothesized that more experienced teachers,
with better classroom management skills at their disposal, experience
less physiological stress when dealing with classroom disruptions.
Contrary to our expectations, we found no buffering effect of teaching
experience on teachers' HR, i.e., more experienced teachers did not show
lower mean standardized HR during the stressful teaching phase than less
experienced teachers. Rather, at least descriptively, we observed the
opposite trend. There are several possible explanations for this
finding. First, teaching experience is inherently confounded with age
(the two variables correlated at \(r = .94\) in our sample), and age has
been shown to affect indicators of cardiovascular reactivity in various
ways \citep{uchino2010older}. However, to avoid this kind of confounding
influence, we had used not raw BPM but rather standardized mean HR for
all our analyses, thus controlling at least for inter-individual
differences in mean HR. Second, as research on teacher
professionalization has repeatedly shown, professional experience is not
a guarantee for higher professional knowledge and skills
\citep{kirschner2016professionswissen}. Rather, honing skills from
professional experience necessitates a deliberate practice of choosing
to improve, learning through experience, and integrating new knowledge
into future performances \citep{dunn1999deliberate}. Thus, rather than
professional experience alone, more direct assessments of classroom
management skills, such as objective behavior-based tests, would be a
better indicator of expertise that future studies could explore.
Finally, and most importantly, the highly controlled teaching situation
that we created in the lab might not have provided sufficient
resemblance to the expert teachers' working conditions to let them
effectively use their coping resources. In other words, since the
situation was unfamiliar to both experienced and unexperienced teachers,
their stress levels might have been more similar than they would have
been in a more authentic classroom setting.

While we did not find a buffering effect of teaching experience on mean
HR during teaching, we did, however, find a less steep HR increase in
more experienced, compared to less experienced teachers during the
pre-teaching phase (\(\beta = -.27\)), i.e., in preparation for the
micro-teaching unit. This finding supports the idea that, even though
teaching experience guarantees neither superior expertise nor stress
resistance, the habits and routines formed by experienced teachers may
at least lead to lower arousal levels (e.g., experienced as feeling less
nervous and tense) when they anticipate potentially stressful teaching
situations.

An interesting observation beyond our hypotheses was that teaching
experience was predictive of HR differences, not during teaching, but in
the interview phase: compared to less experienced teachers, more
experienced teachers showed a higher mean standardized HR
(\(\beta = .24\)) and, thus, probably experienced higher levels of
physiological stress during the SRI. One possible explanation for this
finding could again be the higher age of the more experienced teachers,
along with slower recovery from stress in older teachers. For instance,
\citet{ritvanen2006responses} observed that, compared to their younger
colleagues, older teachers did not experience a decrease in their HR
during periods of low stress levels, from which they concluded that
recovery from stress was insufficient in the older teachers
\citep{ritvanen2006responses}. Alternatively, the finding could also be
attributed to the fact that less experienced teachers, due to their
ongoing or only recently concluded training, may have been more
accustomed to reflecting on their work and receiving feedback as was the
case during the SRI, whereas, these activities were less routine and
possibly more face-threatening for experienced teachers. Therefore, it
is possible that more experienced teachers found the interview itself to
be more stressful and therefore showed a higher mean standardized HR
during this phase.

With regards to the predictive power of teachers' subjective appraisals
of the classroom disruption during teaching, we, first of all, have to
conclude that our hypotheses were not supported, as neither confidence
appraisal nor disruptiveness appraisal showed any notable correlations
with teachers' means standardized HR or any explanatory power over and
beyond teaching experience. Possibly, teachers' self-reported
appraisals, and their actual physiological stress responses, tap into
quite different phenomena, or at least, quite different aspects of the
multifaceted stress response \citep{kyriacou1978}. In addition, while HR
was assessed online during teaching, self-reported appraisals were given
in retrospect during the SRI, and may be subject to biased (e.g.,
self-serving) reporting or simply an inability to recall one´s immediate
stress reactions.

On the other hand, when controlling for all other factors, teachers who
reported to have perceived the events as more disruptive showed a higher
HR (\(\beta = .25\)) in the phase immediately following the
micro-teaching unit. This finding would be consistent with the idea that
differences in mean HR, as an indicator of the physiological stress
response, can be linked to the cognitive appraisal of stressors.

\subsection{Limitations and future
directions}\label{limitations-and-future-directions}

While the laboratory setting of the study allowed for a controlled
implementation of stressors and high internal validity, it was not an
authentic classroom environment, raising questions about its external
validity. Most importantly, the teacher and their students did not have
a shared history, and only a very thin basis for establishing a positive
teacher-student relationship, which is a core characteristic of
effective classroom management \citep{ruedi2014, beaty2010}. In
addition, the micro-teaching unit was only about 15 minutes long, and
thus much shorter than a regular school lesson, providing less
opportunities for experienced teachers to build up an engaging lesson.
Finally, student behavior was scripted, with classroom disruptions
following the experimental schedule, irrelevant of the behavior of the
teacher. Thus, the setting may have masked effects of teaching
experience by providing too little opportunities of experienced teachers
to demonstrate their true classroom management skills, in particular
regarding the prevention of disruptions. In subsequent studies, it would
therefore be insightful to assess teachers' HR in more authentic
classroom settings over a longer period of time (e.g., days, weeks, or
even months). Extended observation of teachers' HR in authentic
classroom settings could reveal how factors such as student behavior,
teaching methods, or organizational and administrative demands
contribute to fluctuations in physiological arousal, uncovering insights
into the sustained physiological demands of teaching that short-term
studies may overlook. Finally, linking actual teacher behavior to
potential stressors (e.g., classroom disruptions, noise level, etc.)
would offer insights into teacher coping strategies and their links with
physiological indicators of stress.

Another limitation concerns the assessment of teachers' HR. While our
results demonstrate the usefulness of drawing upon easily available HR
data from ubiquitous, low-cost, un-intrusive fitness trackers to
estimate teacher stress, there also are shortcomings of this type of
assessment. First, while fitness trackers typically yield HR data, heart
rate variability (HRV) has been demonstrated to be an even more accurate
indicator of stress \citep{wettstein2020ambulatory}. While standard
fitness trackers did not provide this measure at the time of our data
collection, more recent products do offer this function. Thus, future
studies might consider assessing HRV instead of HR. Second, we did not
record participants' resting HR, which is generally considered an
important baseline for determining inter- and intrapersonal differences
in cardiovascular health and reactivity
\citep{nanchen2018, heneghan2019}. A clean baseline HR requires a
resting phase without physical movement or emotional stress, ideally
fifteen minutes before the beginning of the activity, which is very
difficult to achieve in practice \citep{sammito2015guideline}, e.g.,
when assessing teacher HR before and during teaching. Thus, our study
explored the possibility of substituting baseline HR measurement via
z-standardization within participants. As a result, the absolute
standardized values of each participant must always be interpreted in
the context of the standardization sample, and thus are less
interpretable than individual BPM values together with a baseline HR.
However, for statistical analyses based on the whole sample, the
standardization fulfills the aim of controlling for differences in
individual HR due to, for example, age-related differences. Finally,
depending on the brand and model of fitness trackers used, the precision
of the HR measurement varies. Research on the reliability of our
deployed Fitbit® device has proven that this brand is generally accurate
in controlled settings and for moderate activity levels
\citep{wallen2016accuracy, hajj2023, fuller2020, jo2016}, as in our
study. For example, the Fitbit® fitness tracker has previously shown
good HR measurement accuracy during resting phases
\citep{jo2016, muggeridge2021measurement} and for activities such as
walking, jogging, and running \citep{hajj2023}. At higher exercise
intensities such as cycling, the Fitbit® tracker may underestimate HR
\citep{thomson2019heart, montoye2017comparative, jo2016, jachymek2021}
but is still within an acceptable range according to systematic reviews
\citep{chevance2022accuracy}. Nevertheless, \citet{gagnon2022} stressed
that Fitbit® trackers cannot replace ECG when precision is paramount.
Despite these considerations, the Fitbit® model appears suitable for our
study purposes, as physical strain was moderate.

Furthermore, while we assessed teachers' appraisals of the stressful
classroom disruptions using a SRI in which they could review the exact
situation, these appraisal ratings were still post-hoc self-reports,
which limits the interpretation of our results. One of the main issues
with post-hoc self-reports is that they rely on the teachers' memories
and subjective interpretations of past events, which may be prone to
various biases such as social desirability \citep{razavi2001self} or
recall errors \citep{van2016accuracy}. Moreover, stress is not a fixed
or stable construct; it is a dynamic, constantly evolving affective
response that can vary depending on context, individual disposition, and
prior experiences, making it particularly challenging to pinpoint valid
and reliable process markers for how individuals appraise stress in
real-time \citep{lazarus1990theory}. While SRIs provide a more detailed
and reflective understanding of the stressor in question, the delayed
nature of the response makes it difficult to capture the immediate,
in-the-moment appraisal that occurs when the stressful event actually
takes place.

\subsection{Hands-on advice for using wrist-worn fitness trackers for
research}\label{hands-on-advice-for-using-wrist-worn-fitness-trackers-for-research}

For researchers aiming to use fitness trackers to collect data, there
are practical aspects to consider concerning the design, data
collection, and data analysis phases of research projects \citep[for an
additional overview, see][]{nelson2020guidelines}:

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\tightlist
\item
  Choosing a suitable model:
\end{enumerate}

Before data collection, researchers need to decide which model of
fitness tracker best suits their research question. One important point
to consider is whether the study will be conducted in the laboratory, in
a clinical environment, or under real-world conditions. Conventional
fitness trackers should not be used if the focus is on measurement
accuracy, such as in medical contexts, as they cannot replace the
accuracy of ECG measurements \citep{gagnon2022}. Moreover, researchers
should consider that measurement accuracy also depends on the intensity
of the movements performed by the participants during data collection.
Fitbit® fitness trackers, for example, underestimate HR at higher
exercise intensities such as cycling
\citep{thomson2019heart, montoye2017comparative, jo2016, jachymek2021}.
For reference, the systematic review by \citet{fuller2020} provides a
detailed overview of studies that used wrist-worn fitness trackers
between 2000 and 2019 and discusses their validity and reliability.
Another point to consider is the price, which at the time of writing
ranged between €30 for the cheapest models and up to €1.700 for medical
wristbands. Currently, models assessing HRV in addition to HR are
becoming more and more affordable and widespread. Still, Fitbit® fitness
trackers might be a good choice for teams operating with moderate
budgets or if larger groups of participants need to be tracked at the
same time. Further, before conducting any study, it should be considered
that the data collected with fitness trackers is health data, and
therefore very sensitive. Researchers have to ensure that their chosen
model of fitness tracker allows them to collect and store data in line
with relevant ethical and legal requirements, for example, guaranteeing
participants' anonymity and secure data storage.

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\setcounter{enumi}{1}
\tightlist
\item
  Operating the fitness tracker:
\end{enumerate}

In planning the operation of their chosen model of fitness tracker,
researchers need to specify the circumference and attachment of the
wrist band and the placement of the fitness tracker on participants. In
particular, researchers conducting studies with children should take
into account their smaller wrist size. When putting on a fitness
tracker, attention must also be paid to whether it is attached to the
dominant or non-dominant wrist, as this can influence HR measurements.
Different models of fitness trackers need to be placed differently and
in line with the manufacturer's instructions. It is also important to
check that the battery is fully charged each time, that the latest
software version is loaded, and that the fitness tracker has been
synchronized before recording data to avoid unnecessary loss of data.
Finally, if researchers want to accurately investigate parameters during
specific time intervals, such as HR during lessons versus breaks, it is
crucial to synchronize the fitness tracker with other time-keeping
devices, such as watches. This synchronization allows researchers to
precisely determine the onset and offset of particular activities or
intervals of interest. By aligning the recorded data with specific time
frames, researchers can ensure that the physiological measurements, such
as HR, are accurately associated with the corresponding periods of
interest. This process enhances the validity and reliability of the data
analysis, enabling a more precise examination of variations in
physiological responses across different time intervals.

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\setcounter{enumi}{2}
\tightlist
\item
  Extracting and analyzing fitness tracker data:
\end{enumerate}

As far as the procedure for processing the data is concerned,
researchers should ensure that the raw data of the physiological
measurements are available for further analysis. For the Fitbit® HR
measurements, for example, the raw data can be downloaded from a website
in the form of .csv files. However, these must be downloaded as soon as
possible after data collection, as some platforms automatically delete
or archive older data files after a certain period due to policies
regarding data storage and retention. This can result in loss of access
to critical data. Additionally, ensuring that data is collected at the
intended sampling rate is crucial for accurate analysis. For instance,
while our fitness tracker was designed to record HR every 1-5 seconds,
we occasionally observed recordings only every 15 seconds, possibly due
to participant movement and tracker placement.

\subsection{Conclusion}\label{conclusion}

This study investigated whether HR data collected from teacher-worn
fitness trackers are suitable for exploring links between HR, subjective
stressor appraisal, and individual teaching experience, to achieve a
more profound comprehension of teacher stress. Results suggest that the
widespread availability of HR data from wearable fitness trackers,
moving ``from heartbeat to data'', presents opportunities both to
teachers for self-monitoring stress levels, and to researchers for
assessing physiological indicators of stress. For example, using fitness
trackers could enable teachers to strengthen their self-awareness in
stressful situations and allow for early self-intervention such as
mindfulness techniques (e.g., deep breathing or body scans)
{[}agyapong2023interventions{]}. Integrating fitness trackers into
teacher training and everyday practice could offer an affordable and
practical method for assessing and managing teacher stress. In teacher
training as well as in research, triangulating data from fitness
trackers, lesson videos, and interviews could provide teachers with
insights into their own stress management, and foster the implementation
of effective stress and classroom management strategies. Taken together,
our findings cater to \citet{wettstein2021} call for the use of
ambulatory assessment methods, particularly in the context of classroom
disruptions, for gaining a deeper understanding of teacher stress and
its impact on both psychological and physiological variables.

In summary, our study contributes to the understanding of stress in
educational settings and underscores the potential of wearable fitness
trackers in advancing research on teacher well-being. By harnessing the
power of wearable technology, we can provide teachers with the tools
needed to better understand and manage their stress, ultimately
enhancing their overall well-being.
